---
title: "Analysis of price reductions"
author: "Prateek"
date: "2025-01-04"
categories: [logistic regression, probability, EDA]
toc: true
draft: true
description: Modelling and analysing probability of house sale price reductions
---

## Introduction
In real estate universe, often listings on the market experience updates due to either changes to property specific information or due to an influence of markets. <br/>
The market influences prices of listings in major way and can lead to price increase or decrease post original advertisement.<br/>
Of interest to a real investment company are price revisions that are < 0, i.e. where prices of listings decrease. Such properties can offer immediate improvement to rental yield investment cases and shorter times to close due to comparatively lower demand.

The current analysis aspires to be able to understand where and based on what factors could price changes have happened using a historic listings dataset. This is to validate the analysis against subject matter knowledge and also expand the current understanding about price revision behaviours in the market. <br/>
A better understanding of factors associated with price reductions can help drive strategic decisions e.g. if prices are more likely to be reduced at a particular location in recent data then the market may be experiencing a contraction there.

The analysis also extends to prediction of probability of price reduction based on most informative inputs, to be able to have a mechanism where operations teams can prioritise negotiations and due diligence for such properties to benefit from the possibility of their weaker market demand.

## Data
The [data](https://raw.githubusercontent.com/prteek/regression_room/refs/heads/main/posts/price_revision/terminal_revisions.csv) is derived using 2 different data files [listings.csv](https://raw.githubusercontent.com/prteek/regression_room/refs/heads/main/posts/price_revision/listings.csv) and [revisions.csv](https://raw.githubusercontent.com/prteek/regression_room/refs/heads/main/posts/price_revision/revisions.csv). <br/>
These files individually contain basic attributes of properties and revised prices (both increased in decreased).

These are cleaned up for sensible choices of attributes in listing data and price revisions and only negative price revisions for listings are retained. There are no missing values and the data values are checked for the suitability of application. The script to cleanup and prepare data can be found [here](https://github.com/prteek/regression_room/blob/main/posts/price_revision/data_cleanup.R).

```{r}
#| echo: false
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(knitr))
library(patchwork)
library(this.path)
setwd(here())
options(width=200)
revisions <- read_csv('terminal_revisions.csv', show_col_types=F) %>%
                mutate(bedroom_count=as.factor(bedroom_count))
kable(head(revisions))
```
The columns are self explanatory and the only thing worth noting is that prices are in GBP (£) and that location refers to a post town in UK.

We've converted *bedroom_count* to a qualitative type, since bedroom count has a highly non linear effect on prices and other attributes of property (1->2 ~= 2 -> 3). <br/>
Additionally, there are only a few different bedroom counts in the data and it could be analysed without treating this information as quantitative.

### EDA

There are unknown values in *tenure* which should likely be either of the other 2 categories. The proportion of these listings is about 8% in the data so we cannot drop these listings since their share is quite large. <br/>
Additionally, most often *flats* are observed to be leasehold (which is sensible given that land is not definitively owned) and *houses* are observed to be freehold. Thus suggesting a strong correlation between *tenure* and *property_type*, which means we can drop this column from analysis and use just the *property_type* for both pieces of information.

```{r}
#| echo: false
kable(revisions %>% group_by(tenure, property_type) %>% summarise(counts=n(), .groups='keep'))

```

We do not find a clear relationship between *asking_price* and 5% reduction in prices. It is still possible for there to be some relationship but it is perhaps conditional on *location* or *property_type* etc.

```{r}
#| echo: false
p1 <- ggplot(revisions) + geom_jitter(aes(x=asking_price, y=is_5percent_reduced), height=0.02, width=0.1, size=0.1) + theme_minimal()
p2 <- ggplot(revisions) + geom_histogram(aes(x=asking_price, fill=is_5percent_reduced), alpha=0.5, bins=25) + theme_minimal()
p1|p2

```

There are locations which have had a higher share of properties reduced and this is encouraging as conditional on location effect other variables may be different.

```{r}
#| echo: false
ggplot(revisions %>%
        group_by(location) %>%
        mutate(prop_yes=sum(is_5percent_reduced == "yes") / n()) %>%
        ungroup(),
        aes(x = reorder(location, prop_yes), fill = is_5percent_reduced)
 ) +
  geom_bar(position = "fill", alpha=0.5) +  # "fill" shows proportions
  labs(title = "Location vs reductions", y = "Proportion") +
  coord_flip() +
  theme_minimal()

```